DP Problems of Interest

### 2012-11-12 docs/roadmap/draft.tex
Merged all the documents into the draft. Now we need to do heavy cleanup.

\input{style.sty}

\title{Draft}
\begin{document}
\maketitle
\pagestyle{headings}
\setcounter{tocdepth}{2} \tableofcontents

% ------------------------------------------------------------------------------------------------
\newpage
\section{Planning/draft}
\subsubsection*{Contributions}\ul
\item Identify DP classes and give a generalization of their shape/dependency graph
\item Provide state of the art/on par parallel implementations for all these classes
\item Generalize/simplify expression by using a DSL to implement these problems
\item Provide multi-platform support (CPU/GPU/FPGA)
\ule

\subsubsection*{Todo @TCK}
\begin{verbatim}
1. check the vanilla implementation
2. read ADP legacy parsers
3. presentation for Thursday
4. Propose an IR and code generators related to
Note. We only need to generate code directly for CUDA, no interaction. Look at string templating .
"""xxx $var xxx"""
- Fork LMS on GitHub and cleanup all the CUDA generator

% class problem into categories
% think of hashmap implementation. is it useful ?
- CPU+GPU implementation for serial (NS?) problems larger than device memory
- optimizations for in-memory computations

1. restriction vs ADP: we give only the best answer, do we care about semi-optimal ones?
2. separate initialization (base cases, atoms) and processing (rules, recurrences) => less cases to handle (no if)
3. use sub-cell parallelism (seen in 2 papers already)

@@

% --------------------------------------------------------------------------------
\input{problems}
\input{benchmarks}

% --------------------------------------------------------------------------------
\bibliographystyle{plain}
\bibliography{bibliography.bib}
\end{document}

>>>>

\documentclass[11pt]{article}
\input{inc/style.sty}

\title{DynaProg for Scala}
\subtitle{A Scala DSL for Dynamic Programming on GPU and FPGA}
\begin{document}
%\maketitle
\shorttitle

\subsection*{Abstract}
Dynamic programming (solving a problem by leveraging subproblems solutions) is a common pattern used in various domains. Yet underlying matrix recurrences are non-trivial and error prone. Additionally, domain experts might not have the skills to make an efficient parallel implementation. In this paper, we leverage the Scala / LMS infrastructure to implement a domain specific language (DSL) for dynamic programming on heterogeneous platforms, which allows to write concise and efficient programs. Our contributions are: \ul
\item Classification of dynamic programming problems, and generalization of their characteristics % shape, dependency graph
\item State of the art parallel implementation of these classes on GPUs
\item Grammar embedded in Scala (DSL) to express dynamic programming problems efficiently
%\item Normalization of the grammar into efficient productions
\item Code generator to transform a grammar into efficient code for CPU, GPU, and FPGA
\item Systematic approach to generate and process backtracking information
\ule

\pagestyle{headings}
\setcounter{tocdepth}{2} \tableofcontents

% ------------------------------------------------------------------------------------------------
\newpage
\section{Introduction}
\subsection{Dynamic programming}
\subsection{Scala and LMS}

% ------------------------------------------------------------------------------------------------
\input{problems}
\input{benchmarks}

% ------------------------------------------------------------------------------------------------
\newpage
\section{Planning/draft}
\subsubsection*{Roadmap}
\begin{verbatim}
2.11 Come with an IR prototype: data structures and nodes
9.11 GenScala on LMS, JNI working for arbitrary types
16.11 GenCuda + LMS CudaCompiler
23.11 Problem generalization: "cyclic keyword", Zucker problem / CudaLoop optimization
30.11-- Gap due to LMS missing knowledge
 7.12 Benchmarking, grammar analysis
14.12
21.12
28.12 --- holiday ---
 4.01 --- holiday ---
11.01 Writing report: implementation description and plan for future work
18.01 Writing report
\end{verbatim}

\subsubsection*{Todo @TCK}
\begin{verbatim}
4. Propose an IR and code generators related to
Note. We only need to generate code directly for CUDA, no interaction. Look at string templating .
"""xxx $var xxx"""
- Fork LMS on GitHub and cleanup the CUDA generator
- CPU+GPU implementation for serial (NS?) problems larger than device memory
- optimizations for in-memory computations

1. restriction vs ADP: we give only the best answer, do we care about semi-optimal ones?
2. separate initialization (base cases, atoms) and processing (rules, recurrences) => less cases to handle (no if)
3. use sub-cell parallelism (seen in 2 papers already) => does not improve

@@

\begin{verbatim}
ideas.txt
---------------------------------------------------
Plan:
1. Define the main problem in focus and its class
   - Dimension
   - Serial/non-serial
   - Monadic/polyadic
2. Define some other problems that belong to this class or simpler classes
3. See the feasibility for FPGA (possibly helped by CPU)
4. Implement this algorithm in both CUDA and FPGA
   -> specify all parameters so that we can compare both implementations
5. Find a common IR for the problems
6. Write a code generator for both architectures from IR
   -> make sure all the test-DP are compiled well from IR
7. Find an expressive and compact language for the user-facing representation

Translation into C++:
http://bibiserv.cebitec.uni-bielefeld.de/macports/resources/download/

----------------------------------

Ideas for the language:
- we want maximum flexibility
- backtrack can be custom, we may want just to give back indices
- two problems: we care about cost and we care about whole backtrack.
	@DP def Fib(n:Int) = if (n<=2) return 1 else Fib(n-1)+Fib(n-2)
  should be translated into the more complete syntax that allows more flexibility
  => get inspiration from papers to write an internal language

- we may want to keep track of indices but we do not want to give access to the backtrack matrix / cost matrix

Language side:
- define a simple way to represent some DP programs
- define a comprehensive way to define them => micro-syntax
- define the mapping into IR (actually the IR might BE the micro-syntax)
- define optimizations to IR
  - making serializable by using aggregation functions/transformations
  - avoiding the cost matrix by moving it into the wavefront
- scheduling (with memory loads for non-serializable)
- code generation for both platforms
  - possibly choice to use which platform for what part (load/compute @ cpu/gpu/fpga)

----------------------------------

Core function F:
- in: s,t strings
- in: neighbor costs: top, left, top+left
- in: neighbor stats: top, left, top+left
- out: backtrack information
- out: cost(i,j)
- out: stats

To be most efficient, we need to get exact bit-sizes of
- strings' char
- costs: min/max values
- stats: min/max values
- backtrack

=> we might want to parallelize processing within a thread that operate on 32 or 64 bit inputs (to reduce memory accesses)

Some ideas:
- define a way to pack the characters => less memory transfer (i.e. GATC=>4 letters in 1 char)
- operate on some larger word (ex 64 bits) to increase thread locality and reduce memory accesses

- write a kernel that takes stats from left and r

		stats (x)
		 ||
		 vv
stats -> KK -> new_stats (y')
(y)		 || \
		 vv  --+ backtrack info (Bxy)
		 new_stats (x')

and compute the backtracking for its cell (optional as template boolean)

- multi-grain
  + within the 64b-words : multiple letters at once (thread level)
  + group threads into blocks that operate on (block level)
  + kernel that operates one after another (GPU/CPU level)
    => can we catch a stream's event at CPU level to alloc memory to get back results ?
- keep track of
  + stats O(m+n)
  + backtrack O(m*n)

- are we sure we need squares ?
  - rectangles can increase the full usage of all threads during (x-y) runs
  - assuming we put longest word vertically we play hot potato for stats
    left->right => block.x exchanges within block, last thread writes to
    global exchange memory (if short, so that it does not penalize running time)
  - we may go up to running in stripes

- if we put in diagonal-major data in the memory, why not stream all chunks like
  that using a cyclic buffer (?) => no empty slot
  ==> this might be useful for the backtrack information

- use extensively profiling : CUDA profiler

- Robust DP paper provides 2 insights: coalesced access + GPU synchronization
  => can we do a producer-consumer scenario at the block border so that we can
     execute a large diagonal horizontal/vertical swipe made of multiple blocks?
  => multiple swipes at different delays (so that we avoid syncing at every step

          s-->
  	  +----------------/------------+
	t | B1            /             |
	| +------/-------/              | KERNEL1 (keep benefit of t in shared mem)
	| | B2  /                       | delay between b1 and b2 can be 1% of line length
	v +----/------------------------+
	  |                             |
	  |                             | KERNEL2
	Memory organized as
	b1 -  -
	b1 b1 b1
	b1 b1 b1
	.. .. ..
	b1 b1 b1
	b2 b1 b1
	b2 b2 b1
	b2 b2 b2
	.. .. ..

For stats (aggregation of the table): we need to maintain both an horizontal and vertical front
For backtracking we need to maintain the whole table
     .. | . |
	..A | B | min/max: Min(D)=Min(Min(C),Min(B))
	----+---+ sum    : Sum(D)=Sum(C)+Sum(B)-Sum(A)
	..C | D | avg    : sum both #cells and values then divide at appropriate cell
	----+---+

we many not need to store what's the previous cell, however, the previous
cell information (usually limited range) is much more compact than the score(32-64 bits)

Memory usage of CPU:

we can potentially extend the GPU's memory by storing some information in the CPU RAM, which in general is larger.
For large sequences, this technique may apply to problems where one needs to access elements "all the way to the top" and aggregate them in a non trivial (max,min,sum,avg) way.

Problem parameters:
- bit size (alphabet,stats,cost,backtrack)
- dimensionality (2D,3D for the matrix)
  - 3D problem example is Smith Waterman with arbitrary gap weight function.
- Summarizability of the set: for computing the stats of an element, we can limit ourselves to computing values up to a certain value
  -in SW with gap extending penalty, we can summarize easily.
  -in Nussinov, we _need_ to access all the previous row-column elements for the pairwise sum (without a window limit).

user-interface.txt
--------------------------------------
This is how we would like to encode the DP programs into Scala DSL.

I think the best would be to reuse most of the syntactic elements and write it
as an uneducated programmer would write, apply annotations/additions on top of
it or detect that it is DP and then converge to our system.

We need to keep in mind that we need both to express the backtracking and the scoring at the same time

Let's build some examples

Smith-Waterman (from problems of interest)
--------------------------------------------

trait SWOps {
	type T
	type Cost = Int
	type Backtrack = (i,j) => Direction
	def rightGap(_ :_, c:Char) : [Cost, Backtrack]
}

def myRec = {
	def M[i,j] : Int = max {
		M[i,j-1] + gapRight,
		M[i-1, j] + gapLeft,
		M[i-1,j-1] + delta(S(i),T(j))
	}
}

def SWat(S0: List[Char], T0:List[Char]) : List[Pair[Char,Char]] = {
	def cost(s: Char, t: Char) = if (s==t) 2 else 1 // example cost function

	def DP(S: List[Char], T:List[Char]): [Score,Backtrack,Output[Pair[Char,Char]] = (S,T) match {
		case (_, Nil) => (0, stop)
		case (Nil, _) => (0, stop)
		case Si::Sr, Tj::Tr) =>
			val nw = SWat(Sr,Tr);
			val n  = SWat(Sr,T);
			val w  = SWat(S,Tr);
			val c = cost(Si,Tj);

			if (nw._1+c>=max(n,w)-d) then (nw._1+c,"NW",[Si,Tj])
			else if (n>=w) then (n._1-d, "N",[Si,_])
			else (w._1-d, "W",[_,Tj])
	}
	backtrack(DP(S0,T0))
}

def Nussinov(S0: List[Char]) : List[Pair[Char,Char]] = {
	def cost(s: Char, t: Char) = if (s==t) 2 else 1 // example cost function

	def DP(S: List[Char], T:List[Char]): [Score,Backtrack,Output[Pair[Char,Char]] = (S.zipWithIndex,T.zipWithIndex) match {
		case (_, Nil) => (0, stop)
		case (Nil, _) => (0, stop)
		case (Si::Sr, Tj::Tr) =>
			val nw = SWat(Sr,Tr);
			val n  = SWat(Sr,T);
			val w  = SWat(S,Tr);
			val cols_rows = 
			val c = cost(Si,Tj);

			if (nw._1+c>=max(n,w)-d) then (nw._1+c,"NW",[Si,Tj])
			else if (n>=w) then (n._1-d, "N",[Si,_])
			else (w._1-d, "W",[_,Tj])
	}

	backtrack(DP(S0,S0))
}

- Char, Score, Backtrack, Output are LMS Rep objects so that we can easily detect
  when we need to use dynamic programming.
- backtrack() is a new function that take the Backtrack information and reconstruct
  the desired output backwards

We would unfold this program into two:
1. The dynamic programming that constructs all the solution and that is the core of the algorithm
2. The backtrack function that reconstruct the solution and can be run anywhere (Java, CPU, ...)

URLs.txt
------------------------------------
Some GPU algorithms:
http://hgpu.org/?cat=11

CUDPP libraries (but awfully big resulting binary):
http://code.google.com/p/cudpp/

13 dwarfs:
http://developer.amd.com/afds/assets/presentations/2155_final.pdf

http://tutorials.jenkov.com/java-reflection/fields.html
http://lampwww.epfl.ch/~michelou/scala/scala-reflection.html

XXX: use TypeClass to put a predicate on types
--------------------------------------------

def fun[T: CanTranslateToC](...)
def fun[T](implicit ev:CanTranslateToC[T])

class CanTranslateToC[T] {
  def translate:String
}

implicit def canTranslateInt = new CanTranslateToC[Int] = {
  def translate = "Int"
}

\end{verbatim}

% ------------------------------------------------------------------------------------------------
\newpage
\bibliographystyle{finplain}
\bibliography{bibliography.bib}
\end{document}

### 2012-11-12 docs/roadmap/problems.tex
Merged all the documents into the draft. Now we need to do heavy cleanup.

\item Visualization: \begin{center}\includegraphics[width=8cm]{problems_zucker.pdf}\end{center}
	% source: <<Parallization of dynamic programming recurrences in computational biology>> paper
\item Optimizations: {\color{red} XXX: notice that there are 3 matrices: $W$,$V$ ($VBI$ is part of $V$) that can be expressed using regular matrix, and $F$ that is of different dimension than $W$ and $V$ and requires a special construction (in the wavefront?). We need to find a nice way to encode both its construction and backtrack into the existing framework (implement 1D DP recursively?)}
\ole

@@

XXX: normalization, transforms applied
}

@@

Putting these two formula together, and using most of the device memory available, we obtain the following results with $S_C=4, S_B\footnote{To deal with larger matrices, backtrack data need to be extended.}=4, S_I=1, S_W=0$ and $M_D=2^{30}$:
\begin{center}\includegraphics[width=14cm]{ns_large.pdf}\end{center}

>>>>

\item Visualization: \begin{center}\includegraphics[width=8cm]{inc/zucker.pdf}\end{center}
	% source: <<Parallization of dynamic programming recurrences in computational biology>> paper
\item Optimizations: {\color{red} XXX: notice that there are 3 matrices: $W$,$V$ ($VBI$ is part of $V$) that can be expressed using regular matrix, and $F$ that is of different dimension than $W$ and $V$ and requires a special construction (in the wavefront?). We need to find a nice way to encode both its construction and backtrack into the existing framework (implement 1D DP recursively?)}
\ole

@@

XXX: normalization, transforms applied

XXX: Think how to implement {\tt cyclic} problems and {\tt windowing} properties within the language

\begin{verbatim}
restrict to only 1 max/min (more restrictive than ADP)
for optimization we want to split grammar into binary productions, to do that we need
to have equality on the following formula. We can solve it more easily if functions are linear
\end{verbatim}
\[\min\limits_{i<k_1<k_2<j}\big[ f(i,k_1,k_2,j) \big] \le  \min\limits_{i<k_2<j} \big[ g_2(i,\min\limits_{i<k_1<k_2} \big[ (g_1(i,k_1,k_2  ) \big],k_2,j) \big]  \]

}

@@

Putting these two formula together, and using most of the device memory available, we obtain the following results with $S_C=4, S_B\footnote{To deal with larger matrices, backtrack data need to be extended.}=4, S_I=1, S_W=0$ and $M_D=2^{30}$:
\begin{center}\includegraphics[width=14cm]{inc/ns_large.pdf}\end{center}
